---
title: "New Exercise"
author: "Mark Dunning"
date: "9 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Exercise: Body-Mass Index
- Let's try some vector arithmetic. Here are the weights and heights of five individuals

|Person | Weight (kg) | Height (cm)|
|-------|------------------:|-------------------:|
|*Jo*|70.8|182|
|*Sam*|67.9|179|
|*Charlie*|75.3|169|
|*Frankie*|61.9|175|
|*Alex*|72.4|181|


- Create *weight* and *height* vectors to hold the data in each column using the `c` function. Create a *person* vector and use this vector to name the values in the other two vectors.

##Exercise: Body-Mass Index

1. The body-mass index is given by the formula:- $BMI = (Weight)/(Height^2)$; where Height is given in ***metres***
  + Create a new vector to record this, called `bmi`.

2. Create a new vector `bmi.sorted` where the bmi values are put in increasing numeric order (HINT: look up the help on the `sort` function)
3. The interquartile range (IQR) of a vector is defined as the 75% percentile of the data minus the 25% percentile. Calculate the IQR for our bmi values 
    + check your answer using the `IQR` function

##Answers to exercise

```{r ex1answers1}
weight <- c(65.8,67.9,75.3,61.9,92.4)
height <- c(192,179,169,175,171)
person <- c("Jo","Sam","Charlie","Frankie","Alex")
names(weight) <- names(height) <- person
```

##Answers to exercise

1. The body-mass index is given by the formula:- $BMI = (Weight)/(Height^2)$; where Height is given in ***metres***
  + Create a new vector to record this, called `bmi`.
  
```{r ex1answers2}
bmi <- (weight)/((height/100)^2)
bmi
```


##Answers to exercise

2. Create a new vector `bmi.sorted` where the bmi values are put in increasing numeric order
3. Print the second, third and fourth items in the vector `bmi.sorted`

```{r}
bmi.sorted <- sort(bmi)
bmi.sorted[4] - bmi.sorted[2]
IQR(bmi)
```

## Answers to exercise

Names are usually carried across to the new vector. Sometimes this is what we want (as for `bmi`) but sometimes it is not (when we are comparing values in `bmi.sorted`). We can remove names by setting them to the special NULL value:

```{r}
names(bmi.sorted) <- NULL
bmi.sorted[4] - bmi.sorted[2]
```

