---
title: "Introduction to Solving Biological Problems Using R - Day 2"
author: Mark Dunning, Suraj Menon and Aiora Zabala. Original material by Robert Stojnić,
  Laurent Gatto, Rob Foy, John Davey, Dávid Molnár and Ian Roberts
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_notebook: default
  html_document: default
---

# 5. Report Writing and Wrap-Up


## Creating a markdown file from scratch

***File → New File → R Markdown***

- Choose 'Document' and the default output type (HTML)
- A new tab is created in RStudio
- The header allows you to specify a Page title, author and output type
```
---
title: "Untitled"
author: "Mark Dunning"
date: "22/12/2016"
output: html_document
---
```
##Text formatting
See ***Help*** → ***Markdown Quick Reference*** in RStudio:

- Enclose text in \* to format in *italics*
- Enclose text in \*\* to format in **bold**
- \*\*\* for ***bold italics***
- \` to format like `code`
- \$ to include equations: $e =mc^2$
- \> quoted text: 

>To be or not to be

- See **Help -> Markdown Quick Reference** for more:
    + Adding images
    + Adding web links
    + Tables


## Not quite enough for a reproducible document

- Minimally, you should record what version of R, and the packages you used.
- Use the `sessionInfo()` function
    + e.g. for the version of R I used to make the slides

```{r}
sessionInfo()
```
    

## Defining chunks

- It is not great practice to have one long, continuous R script
- Better to break-up into smaller pieces; '*chunks*'
- You can document each chunk separately
- Easier to catch errors
- The characteristics of each chunk can be modified:
    + You might not want to print the R code for each chunk
    + ...or the output
    + etc.


## Chunk options

Code chunks are encapsulated between backticks. Options for the chunk can be put inside the curly brackets`{...}`

```
'''{r}
my code here...
'''
```

- It's a good idea to name each chunk
    + Easier to track-down errors
- We can display R code, but not run it
    + `eval=FALSE`
- We can run R code, but not display it
    + `echo=FALSE`
    + e.g. setting display options
- Suppress warning messages
    + `warning=FALSE`

    
## Chunk options: eval

- Sometimes we want to format code for display, but not execute; we want to show the code for how we read our data, but want our report to compile quickly

```
'''{r, eval=FALSE}
data <- read.delim("path.to.my.file")
'''
```


## Chunk options: echo

- Might want to load some data from disk
    + e.g. the R object from reading the data in the previous slide
```
'''{r echo=FALSE}
load("mydata.rda")
'''
```
- or your P.I. wants to see your results, but doesn't really want to know about the R code that you used

## Chunk options: results

- Some code or functions might produce lots of output to the screen that we don't need
    + `results='hide'`
```{r results='hide'}
for(i in 1:100) {
  print(i)
  }
```


##Chunk options: message and warning

- Loading an R package will sometimes print messages and / or warnings to the screen
- This is not always helpful in a report:
```
'''{r}
library(DESeq)
'''
```

```{r echo=FALSE}
library(DESeq)
```

##Chunk options: message and warning

- Using `message=FALSE` and `warning=FALSE`
```
'''{r message=FALSE, warning=FALSE}
library(DESeq)
'''
```
- Could also need `suppressPackageStartupMessages`

##Chunk options: cache

- The argument `cache=TRUE` will stop certain chunks from being evaluate if their code does not change
- It speeds-up the compilation of the document
    + we don't want to reload our dataset if we've only made a tiny change downstream
```
'''{r echo=FALSE, cache=TRUE}
load("mydata.rda")
'''
```

## Running R code from the main text

- We can add R code to our main text, which gets evaluated
    + make sure we always have the latest figures, p-values etc
- See example `gene-expression-analysis.Rmd`

# End of Course

## Wrap-up

- **Thanks for your attention**
- Practice, practice, practice
    + ... & persevere
- Need inspiration? R code is freely-available, so read other people's code!
    + Read [blogs](http://www.r-bloggers.com/)
    + Follow the [forums](http://stackoverflow.com/questions/tagged/r)
    + Download [datasets](http://vincentarelbundock.github.io/Rdatasets/datasets.html) to practice with
    + Bookmark some [reference](https://en.wikibooks.org/wiki/R_Programming) guides
    + on twitter @rstudio, @Rbloggers, @RLangTip
    + Attend the [follow-on course](http://training.csx.cam.ac.uk/bioinformatics/event/1800066) on data manipulation and graphics
- Please fill in the feedback form for us to improve the course
    
