---
title: "Exercise 4a"
author: "Your Name"
date: "9 October 2015"
output: pdf_document
---

# Import the ozone dataset into R

```{r}
weather <- read.csv("ozone.csv")
```

# scatter plot two variables. e.g. Solar Radiation against Ozone

```{r}
plot(weather$Solar.R,weather$Ozone)
```

# a histogram. e.g. Temperature

```{r}
hist(weather$Temp)
```

# boxplot of a continuous variable against a categorical variable. e.g. Ozone level per month

```{r}
boxplot(weather$Ozone~weather$Month)
```


# Scatter plot of solar radiation versus ozone level; points coloured in orange with filled circles

```{r}
plot(weather$Solar.R,weather$Ozone,col="orange",pch=16,
     ylab="Ozone level",xlab="Solar Radiation", 
     main="Relationship between ozone level and solar radiation")
```

# Histogram of temperature; density on y axis, coloured purple, broken into 50 bins of equal size

```{r}
hist(weather$Temp,col="purple",xlab="Temperature",
     main="Distribution of Temperature",breaks = 50,freq=FALSE)
```

# Boxplot of Ozone level per-month; different colours for each month

```{r}
boxplot(weather$Ozone~weather$Month,col=rainbow(5))
```

