---
title: "Exercise 2"
author: "Your Name"
date: '`r format(Sys.time(), "%d %b %Y")`'
output: pdf_document
---

## Create a data frame called my.patients using the instructions in the slides. Change the data if you like.

```{r}
age    <- c(50, 21, 35, 45, 28, 31, 42, 33, 57, 62)
weight <- c(70.8, 67.9, 75.3, 61.9, 72.4, 69.9, 63.5, 
71.5, 73.2, 64.8)
firstName <- c("Adam", "Eve", "John", "Mary", "Peter", 
"Paul", "Joanna", "Matthew", "David", "Sally")
secondName <- c("Jones", "Parker", "Evans", "Davis", 
"Baker","Daniels", "Edwards", "Smith", "Roberts", "Wilson")

consent <- c(TRUE, TRUE, FALSE, TRUE, FALSE, FALSE,
FALSE, TRUE, FALSE, TRUE)

sex <- c("Male", "Female", "Male", "Female", "Male", "Male",
"Female", "Male", "Male", "Female")

patients <- data.frame(First_Name = firstName, 
                       Second_Name = secondName, 
                       Full_Name = paste(firstName, secondName), 
                       Sex = factor(sex),
                       Age = age,
                       Weight = weight,
                       Consent = consent,
                       stringsAsFactors = FALSE)

my.patients <- patients
my.patients
```


## Remake your data frame with three new variables: country, continent, and height

+ Make up the data
+ Make country a character vector but continent a factor


## Try the summary function on your data frame. What does it do? How does it treat vectors (numeric, character, logical) and factors? (What does it do for matrices?)

## Use logical indexing to select the following patients: 

+ Patients under 40

```{r}
##PUT YOUR R CODE HERE
```

+ Patients who give consent to share their data

```{r}
##PUT YOUR  R CODE HERE
```

+ Men who weigh as much or more than the average European male (70.8 kg)


```{r}
##PUT YOUR  R CODE HERE
```

